Exploring red wine data by Nika Dogonadze
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.
library(ggplot2)
library(psych)
library(dplyr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data

wqr = read.csv('wineQualityReds.csv') 
```

This report explors dataset about red wine quality, which can be found [here](https://www.google.com/url?q=https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityReds.csv&sa=D&ust=1508499119744000&usg=AFQjCNE00TiuImFciSXwq5rY9JARtyxktA)

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
dim(wqr)

str(wqr)

summary(wqr)
```

This particular dataset contains 13 variables and just under 1,600 observations. The real number of variables is 12, since x is just the row number. 

```{r echo=FALSE}
qplot(x = quality, data = wqr, binwidth=1)
```

The quality of wine ranges from 3 to 8 and seems to have a normal distribution. Most of wines have quality rating of 5 or 6.
Now, Lets look at distribution of other variables

```{r echo=FALSE}
qplot(x = fixed.acidity, data = wqr, binwidth=0.5)
```

This parameter shows how much [tartaric acid](https://en.wikipedia.org/wiki/Tartaric_acid) wine contains, given in grams per litre. As can be seen on the graph it has normal distribution which is slightly skewed to the right. Most wines have fixed acidity arount 6-10 grams per litre.


```{r echo=FALSE}

qplot(x = volatile.acidity, data = wqr, binwidth=0.02)

```

Volative acidity shows measurement of [acetic acid](https://en.wikipedia.org/wiki/Acetic_acid) in wine, which can be a cause of an unpleasant, vinegar taste. It is given in same units as fixed acidity. It look like bimodal distribution with some outliers having too large values.


```{r echo=FALSE, warning=FALSE, message=FALSE}

qplot(x = citric.acid, data = wqr, binwidth=0.05)
nrow(subset(wqr, citric.acid == 0))

```

132 of wines have [citric acid](https://en.wikipedia.org/wiki/Citric_acid) measure of 0. The maximum is 1 grams per litre. 

```{r echo=FALSE}

qplot(x = residual.sugar, data = wqr, binwidth=0.1) + 
  scale_x_log10(breaks=1:16)

```

The measurment of sugar (g per litre) in wine is something I would expect to have a lot of influence on wine quality. However, I have read that wines are only considered sweet after having more than 45 grammes per litre of residual sugar. Maximum in this data is about 15g/litre with most having 4g/litre or less. 


```{r echo=FALSE}

qplot(x = chlorides, data = wqr, binwidth=0.05) + 
  scale_x_log10(breaks = c(0.01, 0.02, 0.03, 0.04, 0.05, 0.1, 0.5))

```

This parameter show amount of salt in g/litre. After scaling x axis to logarithmic scale, it's distirbution looks a lot like normal distribution. I would imagine this also is important part of wine taste.


```{r echo=FALSE}

  qplot(x = free.sulfur.dioxide, data = wqr, binwidth=2)
  qplot(x = total.sulfur.dioxide, data = wqr, binwidth=3)
  qplot(x = sulphates, data = wqr, binwidth=0.05)

```

[sulfur dioxide](https://en.wikipedia.org/wiki/Sulfur_dioxide) can have an influence on microbial growth and win oxidation level. It is not easily detectable in wine taste, unless it is more than 50 parts per million. Sulphates can increase number of sulfur oxides. Sulfur dioxide and suphate levels look heavily skewed to the right, lets see them on logarithmic scale.

```{r echo=FALSE}
  qplot(x = free.sulfur.dioxide, data = wqr, binwidth=0.2) + 
    scale_x_log10(breaks=2**(0:6))

  qplot(x = total.sulfur.dioxide, data = wqr, binwidth=0.2) + 
    scale_x_log10(breaks=2**(0:8))
  
  qplot(x = sulphates, data = wqr, binwidth=0.03) + 
    scale_x_log10(breaks=c(0.4, 0.5, 1, 2))
```

After scaling values to log10 the distributions look normal. Conviniently, we did not have any zeros. 


```{r echo=FALSE}
qplot(x = density, data = wqr, binwidth=0.001)

qplot(x = pH, data = wqr, binwidth=0.05)

qplot(x = alcohol, data = wqr, binwidth=0.3)
```

Density, pH level and alcohol concetrations are probably easiest to taste in wine judging from my personal experience. It would be very interesting to see how they influence wine quality. 


# Univariate Analysis

### What is the structure of your dataset?

The dataset consists of 1599 observations of 12 variables (fixed.acidity, volatile.acidity, citric.acid, residual.sugar, chlorides, free.sulfur.dioxide, total.suluf.dioxide, density, pH, sulphates, alcohol, quality)

All the variables are numeric, with no missing values

### What is/are the main feature(s) of interest in your dataset?

The main features of interest are the alcohol level and quality. I am very curious to see which features are best predictors of wine quality. Depending on the answer to this question it might be possible to build a good predictive model using this dataset. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I think sugar and pH levels will have a lot of influence on wine quality, since those are the things that are easy to taste. 

### Did you create any new variables from existing variables in the dataset?

No, I did not create any new variables from the existing ones.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I have not made any adjustments to the dataset itself, but I scaled sulfur dioxide levels to logarithmic scale to better understand nature of it's distribution.


# Bivariate Plots Section

Lets first look for the pair that have relatively high correlation, hopefully some variables will be highly correlated to quality.

```{r echo=FALSE, Bivariate_Plots}

corr_table <- as.data.frame(as.table(cor(wqr)))

subset(corr_table, abs(Freq) > 0.5 & Var1 != Var2)

```

Unfortunately, there do not seem to be any strongly correlated values with the quality, But Lets look at all the combinations.

```{r echo=FALSE}

pairs.panels(subset(wqr, select = -c(X, sulphates, residual.sugar, chlorides)))

```

The most correlated variable with quality seems to be the alcohol, and volatile acidity. However other variables are still worth exploring of course.

```{r echo=FALSE}
  ggplot(aes(x = alcohol, y = quality), data = wqr) + 
    geom_jitter(alpha=1/5)

  wqr %>%
    group_by(quality) %>%
    summarise(mean_alcohol = mean(alcohol),
              median_alcohol = median(alcohol),
              min_alcohol = min(alcohol),
              max_alcohol = max(alcohol))
```

After adding some jitter to points to fix the overplotting issue, it can be seen from the graph, that higher quality wines generally have higher alcohol concetration. There are quite a few exceptions to this though.

```{r echo=FALSE}
  ggplot(aes(x = volatile.acidity, y = quality), data = wqr) + 
    geom_jitter(alpha = 1/10)

  wqr %>%
    group_by(quality) %>%
    summarise(mean_alcohol = mean(volatile.acidity),
              median_alcohol = median(volatile.acidity),
              min_alcohol = min(volatile.acidity),
              max_alcohol = max(volatile.acidity))
```

Good quality wines have relatively low volatile acidity. The best ones have no more than 0.85 g/litre and the lowest quality wines having as much as 1.58 g/litre.

```{r echo=FALSE}
  ggplot(aes(x = pH, y = quality), data = wqr) + 
    geom_jitter(alpha=1/10)

  ggplot(aes(x = quality, group = quality, y = pH), data = wqr) + 
    geom_boxplot()
```

  Unlike the scatter plot, the boxplot graph shows that wines that are rated with lower quality have higher average pH value. It is showing some outliers too, mostly having quality 5 or 6 which is expected because most of quality values are in that range.
  
  
```{r echo=FALSE}
  ggplot(aes(x = residual.sugar, y = quality), data = wqr) + 
    geom_jitter(alpha = 1/20)

  wqr %>%
    group_by(quality) %>%
    summarise(mean_alcohol = mean(residual.sugar),
              median_alcohol = median(residual.sugar),
              min_alcohol = min(residual.sugar),
              max_alcohol = max(residual.sugar))
  
  ggplot(aes(x = quality, group = quality, y = residual.sugar), data = wqr) + 
    geom_boxplot() + 
    coord_cartesian(ylim=c(1, 4))
```

It is against my intuition that residual sugar content would not have an effect in wine quality. I suppose the possible reason could be that the variation between the sugar levels in this dataset is not noticeable enough for human taste to
have an effect. Or maybe wine experts do not pay attention to sweetness when evaluating wines.

```{r echo=FALSE}
  ggplot(aes(x = alcohol, y = density), data = wqr) + 
    geom_point(alpha = 1/10) + 
    geom_smooth(method = 'lm')
```

It is clear from the plot and the trendline that wines with higher alcohol levels have lower density. That is quite intuitive because of alcohol's low density.

```{r echo=FALSE}
    ggplot(aes(x = free.sulfur.dioxide, y = total.sulfur.dioxide), data = wqr) + 
    geom_point(alpha = 1/10) + 
    geom_smooth(method = 'lm')
```


# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

The quality of the wine does not have a strong correlations with any of other variables. The most correlated
variables to quality turned out to be alcohol and volatile.acidity. 
It seems that better quality wines generally contain more alcohol and have less volatile acidity.


### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

When making scatter plot of alcohol and density I observed that wines with higher alcohol content also have lower
densities, however this relationship is not very strong and there quite a few outliers.


### What was the strongest relationship you found?

The strongest relationship ovserved was between free sulfur dioxide and total sulfur dioxide. Which is not at all surprising as the latter contains the former.

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}
  ggplot(aes(x = alcohol, y = quality, color = density), data = wqr) + 
    geom_jitter()

  ggplot(aes(x = alcohol, y = quality, color = volatile.acidity), data = wqr) + 
    geom_jitter()
```

Higher rated wines have low volatile acidity and contain more alcohol. Almost all the wines that are rated 7 or higher
have either low volatile acidity or high alcohol content.

```{r echo=FALSE}
  wqr$alcohol_binned <- cut(wqr$alcohol, breaks=c(8, 10, 11, 12, 15)) 

  ggplot(aes(x = pH, y = quality), data = wqr) + 
    geom_jitter(alpha = 1/10) + 
    facet_wrap(~alcohol_binned)
``` 

For the alcohol levels between 10%-12% the lower pH level corresponds to higher wine quality assessment. However this does not hold true for wines containing less than 10% or more than 12% alcohol. 


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

At first, despite having some correlation, neither alcohol nor volatile acidity were complete
in accounting for the wine quality score. But when visualising all 3 variables in one plot,
almost all the top rated wine turned out to have good measurement in at least one of those parameters.

### Were there any interesting or surprising interactions between features?

I was very surprised to find that lower pH level indicates higher quality of wine only if the alcohol range is
between 10%-12%.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

ggplot(aes(x = quality), data = wqr, binwidth = 1) +
  geom_bar(fill='blue') + 
  ggtitle('Quality Score Distrobutoin of Red Wines') + 
  xlab('Quality score of red wine') + 
  ylab('Wines count') + 
  scale_x_continuous(breaks = 3:8)

```

### Description One

This is the dsitribution of scores wine experts give to wines. The quality score culd from 0 to 10 inclusive, but
the experts are not very generous with their score, top wines only getting score of 8.


### Plot Two
```{r echo=FALSE, Plot_Two}
  ggplot(aes(x = alcohol, y = density), data = wqr) + 
    geom_point(alpha = 1/5, color = 'blue') + 
    ggtitle('Density Versus Alhocol Percentage in Red Wines') + 
    geom_smooth(method = 'lm', color = 'black') + 
    xlab('% of alcohol in wine (by volume)') + 
    ylab('Density of wine (compared to water)')
```

### Description Two

It is always nice to be surpised by the data, but sometimes it is also pleasant and reassuring to see the expected trends. In this case we see that the more alcohol wine contains the less density it has, which is what one would expect considering the low denisty of the alcohol compared to water.



### Plot Three
```{r echo=FALSE, Plot_Three}
  ggplot(aes(x = alcohol, y = quality, color = volatile.acidity), data = wqr) + 
    geom_jitter() + 
    ggtitle('Quality Score Versus Alcohol Percentage in Red Wines') + 
    ylab('Quality score of wine') + 
    xlab('% of alcohol in wine (by volume)')
```

### Description Three

This graph shows alcohol percentage, volatile acidity in g/litre and quality score of wines. It illustrates that most of high-score wines are on the right side or have dark shade of blue which means high alcohol percentage or low volatile acidity respectively.

------

# Reflection

This dataset of red wines contains information about 1599 different wines tasted and evaluated by wine experts. I stated out by analysing variables individually and then in combinations. I discovered some relationships through visualizations and analysis.

It was very surprising for me that not a single variable had a strong correlation with the wine quality score. Relatively stronger correlations were alcohol percentage and volatile acidity. None of them seemed to be a good indicator of wine quality, but when plotting all those together made it clear that both are important factors for whether or not an expert will like the wine.

The next very interesting step would be to build a model for quality score prediction. It's performance could be less than perfect because of some data limitations. The problem I see is that all the experts probably have a bias towards different tastes. Becaues of this, the generalisation of the wine quality score could be limited by their subjectivity. What I would improve in the data is to include identifiers for experts who rated the wine.
